---
title: "Mixed models"
author: "Filippo Biscarini"
date: "12/31/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Mixed models

Linear mixed models are a very powerful statistical tool that, simply put, allows to mix "fixed" (systematic) and "random" effects in one model equation:  

$$
\mathbf{y} = \mathbf{Xb} + \mathbf{Zu} + \mathbf{e}
$$

- **y**: vector of response (target) variable (observations) $(n,1)$
- **b**: vector of fixed effects $(f,1)$
- **u**: vector of random effects $(r,1)$
- **e**: vector of model residuals $(n,1)$
- **X**: design matrix that relates observations (responses) with fixed effects $(n,f)$
- **Z**: design matrix that relates observations (responses) with random effects $(n,r)$

The *letters* between brackets refer to the vector/matrix dimensions: *n* for the number of records (observations), *f* for the number of fixed effects, *r* for the number of random effects.

This is a **linear additive mixed model**: 

- *linear* because it is a linear combination of variables (a.k.a. "parameters"): **b**, **u** (and *e*)
- *additive* because variables are summed together
- *mixed* because (again) it mixes fixed and random effects

While **fixed effects** are just point estimates of an underlying *true* value, **random effects** are associated with a distribution and, importantly, a **variance component**. 
The total variance of the observations (**y**'s) is partitioned into one component belonging to the random effect **u** and the residual variance:

$$
Var(\mathbf{y}) = \mathbf{K} \sigma_u^2 + \mathbf{R} \sigma_e^2
$$
where:

- **K**: *covariance* (relationship) *matrix* between the observations (for the specific random effect)
- **R** is the residual covariance matrix, usually simplified to $\mathbf{I}$ (identity matrix) under the assumption of independence between residuals


### Mixed models: an illustration in R

We take data on the yield of 41 potato lines (`DT_example` dataset) from different environments (combination of location and year).
There are *185* records.

```{r mixed_model, warning=FALSE, message=FALSE}
library("sommer")
library("tidyverse")
data(DT_example)
DT <- DT_example

head(DT)
```

We first look at the distribution of the quantitative response variable:

```{r}
hist(DT$Yield, xlab = "yield", main="response variable")
```
Then at the distribution of observations (yield) per environemnt:

```{r}
p <- ggplot(DT, aes(x = Name, y = Yield)) + geom_jitter(aes(color=Name)) + facet_wrap(~Env)
p <- p + theme(legend.position = "none", axis.text.x = element_blank())
p
```
### Fitting the mixed model

We now fit a mixed model of the form:

$$
\mathbf{yield} = \mathbf{X} \cdot \mathbf{env} + \mathbf{Z} \cdot \mathbf{variety} + \mathbf{e}
$$

```{r}
fit <-mmer(fixed = Yield ~ Env, 
            random= ~ Name,
            rcov= ~ units,
            data=DT, 
            verbose = FALSE)
```

```{r}
summary(fit)
```

