---
title: "BLUP & GBLUP"
author: "Filippo"
date: "2023-04-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

base_folder = "/home/filippo/Documents/ciampolini/unipisa_2023/bioinformatics_and_biostatistics_training/introduction_to_animal_breeding"

library("sommer")
library("pedigree")
library("tidyverse")
library("data.table")
```

## BLUP

Here we illustrate the animal model with the dog (Braque Français) dataset:

```{r cars}
phenotypes = fread(file.path(base_folder, "data/pheno.dat"))
ped <- fread(file.path(base_folder, "data/pedigree.txt"))
```

Let's create the additive relationship matrix from pedigree data:

```{r pressure, echo=FALSE}
animals <- rep(TRUE,nrow(ped))
makeA(ped, animals)
A <- read.table("A.txt",header=FALSE)
```

```{r}
triA <- A |>
  spread(key = "V2", value = "V3") |>
  dplyr::select(-V1)

tA = t(triA)
triA[upper.tri(triA)] <- tA[upper.tri(tA)]
```

```{r}
vec <- ped$dog_id %in% phenotypes$id
A <- as.matrix(triA[vec,vec])
colnames(A) <- phenotypes$id
rownames(A) <- phenotypes$id
```

```{r}
heatmap(A)
```


#### BLUP model

We use one fifth (1/5) of the data as validation (phenotype set to missing $\rightarrow$ "masked" phenotypes)

```{r}
## sample 1/5 of the phenotypes as validation set (set their phenotype to NA)
y.trn <- dplyr::select(phenotypes, c(id, sex, game))
vv <- y.trn$id %in% sample(y.trn$id,round(nrow(y.trn)/5))
y.trn[vv,"game"] <- NA
y.trn
```

Ready to run the BLUP model!

```{r}
res = mmer(game ~ sex,
           random = ~vsr(id, Gu=A),
           rcov = ~units,
           nIters = 10,
           data = y.trn)
```

#### Predictions

We can now make predictions in the training set:

```{r}
## predictions
ids = filter(y.trn, is.na(game)) |> pull(id)
y_hat = res$U$`u:id`$game
y_hat = as.data.frame(y_hat)
y_hat$id = row.names(y_hat)
y_hat <- y_hat |> inner_join(y.trn, by = "id")


sex_effect = filter(res$Beta, Effect == "sex") |> pull(Estimate)
y_hat <- y_hat |> mutate(y_hat = y_hat + sex_effect*sex)

## model performance in the training set
pred_train <- filter(y_hat, !(id %in% ids))
pred_train <- mutate(pred_train, sex = as.factor(sex))
print(paste("The correlation between predicted and observed phenotypes in the TRAINING SET is: ", cor(pred_train$y_hat, pred_train$game)))
```

```{r plot_training_preds}
ggplot(pred_train, aes(x = y_hat, y = game)) + geom_jitter(aes(color=sex))

## MSE?
mse_train = sqrt(sum((pred_train$y_hat - pred_train$game)^2)/nrow(pred_train))
mse_train/mean(phenotypes$game)
```

We can now make predictions in the **validation set**:

```{r}
## model performance in the validation set
## observations in the validation set
y = filter(phenotypes, id %in% ids) |> dplyr::select(id, game)
pred_val = dplyr::select(y_hat, -game) |> inner_join(y, by = "id")
pred_val <- mutate(pred_val, sex = as.factor(sex))
print(paste("The correlation between predicted and observed phenotypes in the VALIDATION SET is: ", cor(pred_val$y_hat, pred_val$game))) ## correlation between observed and predicted values
```

```{r corr_validation}
ggplot(pred_val, aes(x = y_hat, y = game)) + geom_jitter(aes(color=sex))

## MSE?
mse_val = sqrt(sum((pred_val$y_hat - pred_val$game)^2)/nrow(pred_val))
mse_val/mean(phenotypes$game)
```

## GBLUP

Enter genomics:

```{r marker_data}
base_folder = "/home/filippo/Documents/ciampolini/unipisa_2023/bioinformatics_and_biostatistics_training/introduction_to_animal_breeding"
mrkrs = fread(file.path(base_folder, "data/bra_imp_maf0.05.raw"))

mat <- as.matrix(mrkrs[,-c(1:6)])
```

Let's make the genomic relationship matrix from SNP marker data:

```{r}
K <- A.mat(mat, endelman = FALSE)
colnames(K) <- mrkrs$IID
rownames(K) <- mrkrs$IID
```

```{r}
heatmap(K)
```

SAme approach as before: 1/5 of the data is used for validation (setting phenotypes to missing):

```{r}
## sample 1/5 of the phenotypes as validation set (set their phenotype to NA)
y.trn <- dplyr::select(phenotypes, c(id, sex, game))
vv <- y.trn$id %in% sample(y.trn$id,round(nrow(y.trn)/5))
y.trn[vv,"game"] <- NA
y.trn
```

And now we run the **GBLUP model**:

```{r}
res = mmer(game ~ sex,
           random = ~vsr(id, Gu=K),
           rcov = ~units,
           nIters = 10,
           data = y.trn)
```


```{r}
## predictions
ids = filter(y.trn, is.na(game)) |> pull(id)
y_hat = res$U$`u:id`$game
y_hat = as.data.frame(y_hat)
y_hat$id = row.names(y_hat)
y_hat <- y_hat |> inner_join(y.trn, by = "id")


sex_effect = filter(res$Beta, Effect == "sex") |> pull(Estimate)
y_hat <- y_hat |> mutate(y_hat = y_hat + sex_effect*sex)

## model performance in the training set
pred_train <- filter(y_hat, !(id %in% ids))
pred_train <- mutate(pred_train, sex = as.factor(sex))
cor(pred_train$y_hat, pred_train$game)
```

```{r}
ggplot(pred_train, aes(x = y_hat, y = game)) + geom_jitter(aes(color=sex))

## MSE?
mse_train = sqrt(sum((pred_train$y_hat - pred_train$game)^2)/nrow(pred_train))
mse_train/mean(phenotypes$game)
```


```{r}
## model performance in the validation set
## observations in the validation set
y = filter(phenotypes, id %in% ids) |> dplyr::select(id, game)
pred_val = dplyr::select(y_hat, -game) |> inner_join(y, by = "id")
pred_val <- mutate(pred_val, sex = as.factor(sex))
cor(pred_val$y_hat, pred_val$game) ## correlation between observed and predicted values
```

```{r gblup_validation}
ggplot(pred_val, aes(x = y_hat, y = game)) + geom_jitter(aes(color=sex))

## MSE?
mse_val = sqrt(sum((pred_val$y_hat - pred_val$game)^2)/nrow(pred_val))
mse_val/mean(phenotypes$game)
```
