---
title: "Genomic inbreeding & ROH"
author: "Filippo"
date: "2023-04-15"
output: html_document
---

```{r setup, include=FALSE}
library("tidyverse")
library("data.table")
```

### The data

We use SNP (genomic) data from a sample of *Braque Français dogs*, to calculate **genomic measures of inbreeding**.

```{r marker_data}
base_folder = "/home/filippo/Documents/ciampolini/unipisa_2023/bioinformatics_and_biostatistics_training/introduction_to_animal_breeding"
mrkrs = fread(file.path(base_folder, "data/bra_imp_maf0.05.raw"))

mrkrs[1:12,1:12]
```

## Genomic inbreeding $\text{F}_{\text{HOM}}$

First, we start with $\text{F}_{\text{HOM}}$ which based on single SNP genotype data. Basically, we calculate the following ratio:

(observed homozygous count - expected homozygous count\>) / (total n. of SNPs - expected homozygous count)

This is based on counting of the genotyping data and on the **Hardy-Weinberg equilibrium**: $(p+q)^2$: $p^2$, $q^2$, $2pq$

For each individual we:

-   count the n. of homozygous SNPs (AA or BB)
-   calculate the expected n. of homozygous SNPs ($2pq$)
-   count the total n. of SNPs

```{r}
mat <- as.matrix(mrkrs[,-c(1:6)])
```

```{r}
## total n. of SNPs
nsnps = ncol(mat)
```

#### Calculate allele frequencies

We now caclulate $p$ and $q$, the frequencies of the two SNP alleles, necessary to compute the expected homozygosity counts under Hardy-Weinberg:

```{r}
aa = rowSums(mat == 0)
ab = rowSums(mat == 1)
bb = rowSums(mat == 2)
```

```{r}
tot <- aa+ab+bb
mean(tot) == ncol(mat) ## sanity check
```

```{r}
sd(tot) ## further sanity check (checks to zero)
```

-   $p$: frequency of *A* alleles $\rightarrow$ twice the number of "AA" genotypes, plus once the number of "AB" genotypes, divided by twice the n. of SNPs (two alleles for each SNP)
-   $q$: this is simply the complement to one of $p$ (since the two relative frequencies must sum to 1)

```{r}
p = (2*aa+ab)/(2*nsnps)
q = 1-p
```

#### Calculate observed and expected homozygous SNP

```{r}
obs_hom = rowSums(mat != 1)
obs_het = rowSums(mat == 1)
exp_het = nsnps*(2*p*q)
exp_hom = nsnps-exp_het
```

#### Calculate genomic inbreeding

Now we have all the ingredients to calculate $\text{F}_{\text{HOM}}$:

```{r}
F_hom = (obs_hom - exp_hom)/(nsnps-exp_hom)
```

Or, equivalently:

```{r}
F_is = 1-obs_het/exp_het
```

```{r}
plot(density(F_hom))
```


```{r}
boxplot(F_hom)
```

```{r}
summary(F_hom)
```


## Runs of homozygosity

```{r}
library("detectRUNS")
```
#### Detect runs

We first detect **ROH** in individual dogs:

```{r}
runs <- consecutiveRUNS.run(
  genotypeFile = "../data/braque.ped",
  mapFile = "../data/braque.map",
  minSNP = 20,
  ROHet = FALSE,
  maxGap = 10^6,
  minLengthBps = 5e+5,
  maxOppRun = 2,
  maxMissRun = 2
)
```


```{r}
plot_Runs(runs = filter(runs, chrom == 6))
```

#### Calculate $\text{F}_{\text{ROH}}$

Genome size from: https://www.ncbi.nlm.nih.gov/assembly/GCF_000002285.3/#/st

```{r}
total_genome = 2410976875
sex_chromosome = 123869142
length_autosome = total_genome - sex_chromosome
```

```{r}
dd <- runs |>
  filter(chrom != "39") |> ## remove the sex chromosome
  group_by(id) |>
  summarise(length_roh = sum(lengthBps)) |>
  mutate(f_roh = length_roh/length_autosome)
```

```{r}
print(dd)
```

```{r}
plot(density(dd$f_roh))
```

```{r}
boxplot(dd$f_roh)
```

```{r}
summary(dd$f_roh)
```

### Relationship between $\text{F}_{\text{HOM}}$ and  $\text{F}_{\text{ROH}}$

```{r}
f_hom = data.frame("id" = mrkrs$IID, "f_hom" = F_hom)
dd <- dd |> inner_join(f_hom, by = "id")
```


```{r}
plot(dd$f_hom, dd$f_roh)
```

```{r}
cor(dd$f_roh, dd$f_hom)
```

