---
title: "Heritability"
author: "Filippo"
date: "2023-03-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

base_folder = "/home/filippo/Documents/ciampolini/unipisa_2023/bioinformatics_and_biostatistics_training/introduction_to_animal_breeding"

library("sommer")
library("pedigree")
library("tidyverse")
library("data.table")

data(DT_example)
```

## Estimate heritability

#### The `sommer` R package

In this notebook, we use the `sommer` *R* package to estimate variance components. The `sommer::mmer` solves the MME (mixed model equations).

The multivariate (and by extension the univariate) mixed model has the form:

$$
Y = X\beta + ZU + e
$$

-   $Y$: vector (matrix) of phenotypes
-   $X\beta$: systematic part of the mixed model
-   $ZU$: random part of the mixed model (e.g. genetic effects)
-   $e$: residuals

Below the related variance structure ($X\beta$ is the mean, $V$ is the variance-covariance matrix):

$$
Y \sim MVN(X\beta, V)
$$

$$
V = 
\begin{bmatrix}
Z_1K\sigma_{g1}^2Z'_i + H\sigma_{e1}^2 & \ldots & Z_1K\sigma_{g1,t}Z'_t + H\sigma_{e1,t} \\
\vdots & \ddots & \vdots \\
Z_1K\sigma_{g1,t}Z'_t + H\sigma_{e1,t} & \ldots & Z_tK\sigma_{gt,t}^2Z'_t + H\sigma_{et,t}^2
\end{bmatrix}
$$

where $i = 1 \ldots t$ indicates the number of traits, $K$ is the covariance matrix (kinship) for the $k_{th}$ random effect, and $H=I$ is an identity (or partial identity) matrix for the residual term.

`sommer` provides kernels to estimate additive (A.mat), dominance (D.mat), and epistatic (E.mat) relationship matrices.

**Heritability** is one of the most popular parameters among the breeding and genetics communities because of the insight it provides in the inheritance of the trait and potential selection response.
Heritability is usually estimated as narrow sense ($h^2$; only additive variance in the numerator $σ^2_A$), and broad sense ($H^2$; all genetic variance in the numerator $σ^2_G$).

### Illustration

We first use a built-in example dataset: 41 potato lines evaluated in 3 environments in an RCBD design

```{r example}
dt_2013 = DT_example |>
  filter(Block == "CA.2013.1")
```

We need to subset the relationship matrix to make it match the data subset we are using (one year):

```{r}
vec <- colnames(A_example) %in% dt_2013$Name
A <- A_example[vec,vec]
dt_2013$Name <- as.character(dt_2013$Name)
```

Now we can run the estimation of variance components:

```{r}
res = mmer(Weight ~ 1,
           random = ~vsr(Name, Gu=A),
           rcov = ~units,
           nIters = 10,
           data = dt_2013)
```

The **variance components**:

```{r varcomp}
summary(res)$varcomp
```

And the **estimated heritability** ($h^2$):

```{r}
vpredict(res, h2 ~ (V1)/(V1+V2))
```

### Dog data

We now use real data from a dog population (Braque Français):

```{r}
phenotypes = fread(file.path(base_folder, "data/pheno.dat"))
```

```{r}
ped <- fread(file.path(base_folder, "data/pedigree.txt"))

animals <- rep(TRUE,nrow(ped))
makeA(ped, animals)
A <- read.table("A.txt",header=FALSE)
```

We need to transform the kinship matrix from the long format to a symmetric matrix:

```{r}
triA <- A |>
  spread(key = "V2", value = "V3") |>
  dplyr::select(-V1)

tA = t(triA)
triA[upper.tri(triA)] <- tA[upper.tri(tA)]
```

Now we ensure correspondance between the phenotype data and the kinship matrix:

```{r example}
vec <- ped$dog_id %in% phenotypes$id
A <- as.matrix(triA[vec,vec])
colnames(A) <- phenotypes$id
rownames(A) <- phenotypes$id
```

And we are ready to go to estimate heritability for one of the traits in the phenotypic dataset (we pick *game*):

```{r}
res = mmer(game ~ 1,
           random = ~vsr(id, Gu=A),
           rcov = ~units,
           nIters = 10,
           data = phenotypes)
```

The **variance components**:

```{r varcomp}
summary(res)$varcomp
```

And the estimated heritability:

```{r}
vpredict(res, h2 ~ (V1)/(V1+V2))
```

### Exercise

Now it's your turn to pick another trait from the file of phenotypes and estimate the heritability:

```{r}
## your code here
```

