---
title: "pedigree relationship matrix"
author: "Filippo"
date: "2023-03-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library("knitr")
library("pedigree")
library("tidyverse")
library("data.table")
```

## Pedigree data - genealogies

We use pedigree data on a small sample from a dog breed (Braque Français):⎄

```{r pedigree_data}
base_folder = "/home/filippo/Documents/ciampolini/unipisa_2023/bioinformatics_and_biostatistics_training/introduction_to_animal_breeding"
ped <- fread(file.path(base_folder, "data/pedigree.txt"))

head(ped)
dim(ped)
```

### Pedigree genetic relationship matrix

We first define a synthetic dataset as example:

```{r}
id <- 1:6
dam <- c(0,0,1,1,4,4)
sire <- c(0,0,2,2,3,5)
pedex <- data.frame(id,dam,sire)
```

Then we write a simple function (not optimised!) to implement the **recursive (tabular) method** for the calculation of the kinship matrix:

```{r}
crA <- function(ped) {
  
  n = dim(ped)[1];
  A = diag(1,n);
  p = ped[,2];
  m = ped[,3];
  
  for(i in 1:n) {        ## browse rows
    for(j in i:n) {  ## column from diagonal
      if(i==j) {          ## if diagonal element
        if(p[j] > 0) {      ## known sire
          if(m[j] > 0) {    ## known dam
            A[i,j] <- A[i,j]+A[p[i],m[i]]/2     ## add inbreeding if present
          }
        }
      } else  {           ## if off-diagonal element
        if(p[j] > 0) {
          A[i,j] < -A[i,j]+A[i,p[j]]/2;         ## add sire contribution
        }
        if(m[j]>0) {
          A[i,j] <- A[i,j]+A[i,m[j]]/2;         ## add dam contribution
        }
        A[j,i] <- A[i,j];     ## A matrix is symmetric
      }
    }
  }
  A;
}
```

Finally, we apply it:

```{r}
crA(pedex)
```

We now use the *R* package `pedigree` to handle the dog pedigree data that we have loaded above (many other options available in R and other software packages). The function `makeA` is used to build the genetic relationship matrix.

```{r}
animals <- rep(TRUE,nrow(ped))
makeA(ped, animals)
A <- read.table("A.txt",header=FALSE)
kable(head(A)) # A matrix (long format)
```

```{r}
A %>%
  filter(V1==V2) %>%
  summarize(mean(V3),sd(V3),min(V3),max(V3)) |>
  kable()
```

```{r}
A %>%
  filter(V1==V2) %>%
  filter(V3>1) %>%
  summarize(n(),min(V3)) |>
  kable()
```

### Generations and inbreeding

Here's a way to obtain the generation number for each dog in the pedigree:

```{r}
gens <- countGen(ped)
gens <- gens[animals]
gens
```

Now we select the kinship coefficients of each animal with itself (**pedigree inbreeding**), and we calculate the average inbreeding by generation:

```{r}
Fped <- A %>%
  filter(V1==V2)

Fped <- cbind.data.frame(Fped,gens)

dd <- Fped %>%
  group_by(gens) %>%
  summarize(avg_inbreeding = mean(V3))

print(dd)
```

We can also plot inbreeding as a function of the number of generations:

```{r}
# ggplot(dd, aes(x = gens, y = avg_inbreeding)) + geom_bar(stat="identity")
ggplot(dd, aes(x = gens, y = avg_inbreeding)) + geom_point() + geom_smooth(method=lm,se = FALSE)
```


```{r}
mA <- A %>%
  filter(V1!=V2) %>%
  dplyr::rename(V2=V1, V1=V2)

mA <- bind_rows(A,mA)
```

```{r}
p <- ggplot(mA, aes(V1,V2))
p <- p + geom_tile(aes(fill=V3)) + xlab("id1") + ylab("id2")
# p <- p + scale_x_continuous(breaks=unique(mK$id1),labels=ids$id)
p <- p + theme(axis.text.x = element_text(angle=90, hjust = 1,vjust=1,size = 8))
p <- p +  scale_fill_gradient2(low = "white", 
                               mid = "yellow", 
                               high = "red", 
                               midpoint = 0.5)
p <- p + guides(fill=guide_legend(title="A"))
print(p)

# ggsave(plot = p, device = "pdf", filename = "matrix_A.pdf", width = 10, height = 10)
```

```{r}
calcInbreeding(ped)
```
